Bare minimum: order theory

### 2025-09-21 BareMinima/OrderTheory/BareMinimumOrderTheory.tex
OrderTheory: final review

Order theory studies partially ordered sets (posets): sets equipped with a notion of order. These structures are pervasive in mathematics as they play a role in logic, computer science, topology, measure theory, group theory, and beyond. Even number types (i.e. integers, rational and reals) can be characterized purely by their order structure. Some physical concepts (e.g. part-whole inclusion) are also naturally captured with order theory.

@@

\begin{defn}[Subposet]
	Given a poset $(X, \leq)$, a \textbf{subposet} is a poset $(Y, \leq_Y)$ where $Y \subset X$ and $\leq_Y = \leq |_Y$.
\end{defn}

@@

\begin{defn}[Linear order]
	A \textbf{linear order}, \marginleft{Linear order,\\ chain} or \textbf{total order}, is a partial order where every pair is ordered. That is, given $a, b \in X$ either $a \leq b$ or $b \leq a$. A \textbf{chain} typically refers to a subset of a partial order that is linearly ordered.
\end{defn}

\begin{defn}[Maximal subsets]
	A linearly \marginleft{Maximal subset} ordered subset (i.e. a chain) is \textbf{maximal} if it is not a scrict subset of another linearly ordered subset.
\end{defn}

\begin{prop}[Kuratowski's Lemma or Hausdorff maximal principle]
	Every linearly ordered subset is the subset of some maximal linearly ordered subset.
\end{prop}

@@

\begin{defn}[Discrete order]
	A \textbf{discrete order} \marginleft{Discrete order} is an order in which every element that has a predecessor/successor has an immediate predecessor/successor.
\end{defn}

Note that discrete order may also mean an order in which every element is only related to itself (i.e. the partial order is an equivalence relation).

@@

\begin{defn}[Lattice properties]
	We define\marginleft{Complete, distributive, complemented} the following additional properties on lattices:
	\begin{description}
		\item[Completeness] A lattice is \textbf{complete} if all subsets have joins and meet, not just finite ones.
		\item[Modularity] A lattice is \textbf{modular} if $a \leq b$ implies $a \vee (c \wedge b) = (a \vee c) \wedge b$
		\item[Distributivity] A lattice is \textbf{distributive} if it satisfies
		\begin{align*}
			a \vee (b \wedge c) = (a \vee b) \wedge (a \vee c) \\
			a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge c).
		\end{align*}
		A complete lattice is \textbf{infinitely distributive} if the same holds over infinite operations.
		\item[Complemented] A bounded lattice is \textbf{complemented} if every element has a complement. It is \textbf{uniquely complemented} if the complement is unique.
		\item[Orthocomplemented] A bounded lattice is \textbf{orthocomplemented} if for every element $a$ there is a complement $a^{\bot}$ such that $a^{\bot\bot} = a$ and if $a \leq b$ then $b^{\bot} \leq a^{\bot}$.
		\item[Orthomodular] A lattice is \textbf{orthomodular} if it is orthocomplemented and if $a \leq b$ implies $a \vee (a^{\bot} \wedge b) = (a \vee a^{\bot}) \wedge b$ (i.e. modularity holds at least of orthocomplements)
	\end{description}
\end{defn}
\marginright{
	$$
	\def\arl{\ar@{-}}
	\xymatrix@R=7pt@C=4pt{
		& M_3 & \\
		 & \top\arl[dl]\arl[d]\arl[dr] & \\
		a\arl[dr] & b\arl[d] & c\arl[dl] \\
		& \bot \\
	}
	\xymatrix@R=2pt@C=4pt{
		& N_5 & \\
		& \top\arl[dl]\arl[ddr] & \\
		a\arl[dd] &  & \\
		 &  & c\arl[ddl] \\
		b\arl[dr] &  &  \\
		& \bot \\
}
	$$
}
\begin{prop}
	Every distributive lattice is a modular. Every complete lattice is bounded.
\end{prop}

\begin{prop}
	A lattice is distributive if an only if it does not contain an $M_3$ (diamond) or $N_5$ (pentagon) sublattice.
\end{prop}

@@

\begin{remark}
	The completion is constructed by finding pairs $(A, B)$ of subsets $A, B \subseteq X$ such that $A$ is the set of lower bounds for $B$ and $B$ is the set of upper bounds of $A$. For the rationals $\mathbb{Q}$, these pairs correspond to Dedekind cuts, and the completion of $\mathbb{Q}$ as a lattice is the lattice of the reals $\mathbb{R}$.
\end{remark}

@@

\begin{defn}[Boolean algebra]
	An algebraic \marginleft{Boolean \\ algebra} structure $(X, \vee, \wedge, \neg, \bot, \top)$ is a \textbf{Boolean algebra} if it satisfies the following properties:
	\begin{description}
		\item[Lattice] $(X, \vee, \wedge)$ is a lattice
		\item[Distributivity] $a \vee (b \wedge c) = (a \vee b) \wedge (a \vee b)$
and $a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge b)$
		\item[Identity] $a \vee \bot = a$
		and $a \wedge \top = a$
		\item[Complementation] $a \vee \neg a = \top$
		and $a \wedge \neg a = \bot$
	\end{description}
\end{defn}

@@

\begin{prop}[Stone's representation theorem]
	Every distributive lattice is order isomorphic to a lattice of subsets ordered by inclusion.
\end{prop}

>>>>

Order theory studies partially ordered sets (posets): sets equipped with a notion of order. These structures are pervasive in mathematics as they play a role in logic, computer science, topology, measure theory, group theory, and beyond. Even number types (i.e. integers, rationals and reals) can be characterized purely by their order structure. Some physical concepts (e.g. part-whole inclusion) are also naturally captured with order theory.

@@

\begin{defn}[Subposet]
	Given a poset $(X, \leq)$, a \textbf{subposet} is a poset $(Y, \leq_Y)$ where $Y \subseteq X$ and $\leq_Y = \leq |_{Y \times Y}$.
\end{defn}

@@

\begin{defn}[Linear order]
	A \textbf{linear order}, \marginleft{Linear order,\\ chain, \\anti-chain} or \textbf{total order}, is a partial order where every pair of elements is comparable. That is, given $a, b \in X$ either $a \leq b$ or $b \leq a$. A \textbf{chain} typically refers to a subset of a partial order that is linearly ordered. An \textbf{anti-chain} is a subset of a partial order in which elements are only comparable to themselves.
\end{defn}

\begin{defn}[Maximal subsets]
	A linearly \marginleft{Maximal subset} ordered subset (i.e. a chain) is \textbf{maximal} if it is not a strict subset of another linearly ordered subset.
\end{defn}

\begin{prop}[Kuratowski's Lemma or Hausdorff maximal principle]
	Every chain is the subset of some maximal chain (i.e. maximal linearly ordered subset).
\end{prop}

@@

\begin{defn}[Discrete order]
	A \textbf{discrete order} \marginleft{Discrete order} is an order in which every element that has a predecessor/successor has an immediate predecessor/successor. A \textbf{discrete linear order} is an order that is both linear and discrete. Be aware that discrete order often refers to an anti-chain.
\end{defn}

@@

\begin{defn}[Lattice properties]
	We define\marginleft{Complete, distributive, complemented} the following additional properties on lattices:
	\begin{description}
		\item[Completeness] A lattice is \textbf{complete} if all subsets have joins and meet, not just finite ones.
		\item[Modularity] A lattice is \textbf{modular} if $a \leq b$ implies $a \vee (c \wedge b) = (a \vee c) \wedge b$
		\item[Distributivity] A lattice is \textbf{distributive} if it satisfies
		\begin{align*}
			a \vee (b \wedge c) = (a \vee b) \wedge (a \vee c) \\
			a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge c).
		\end{align*}
		A complete lattice is \textbf{infinitely distributive}, or \textbf{completely distributive} if the same holds over infinite operations.
		\item[Complemented] A bounded lattice is \textbf{complemented} if every element has a complement. It is \textbf{uniquely complemented} if the complement is unique.
		\item[Orthocomplemented] A bounded lattice is \textbf{orthocomplemented} if for every element $a$ there is a complement $a^{\bot}$ such that $a^{\bot\bot} = a$ and if $a \leq b$ then $b^{\bot} \leq a^{\bot}$.
		\item[Orthomodular] A lattice is \textbf{orthomodular} if it is orthocomplemented and if $a \leq b$ implies $a \vee (a^{\bot} \wedge b) = (a \vee a^{\bot}) \wedge b$ (i.e. modularity holds at least of orthocomplements)
	\end{description}
\end{defn}
\marginright{
	$$
	\def\arl{\ar@{-}}
	\xymatrix@R=7pt@C=4pt{
		& M_3 & \\
		 & \top\arl[dl]\arl[d]\arl[dr] & \\
		a\arl[dr] & b\arl[d] & c\arl[dl] \\
		& \bot \\
	}
	\xymatrix@R=2pt@C=4pt{
		& N_5 & \\
		& \top\arl[dl]\arl[ddr] & \\
		a\arl[dd] &  & \\
		 &  & c\arl[ddl] \\
		b\arl[dr] &  &  \\
		& \bot \\
}
	$$
}
\begin{prop}
	Every distributive lattice is a modular. Every complete lattice is bounded.
\end{prop}

\begin{prop}
	A lattice is distributive if and only if it does not contain an $M_3$ (diamond) or $N_5$ (pentagon) sublattice.
\end{prop}

@@

\begin{remark}
	The Dedekind-MacNeille completion is constructed by finding pairs $(A, B)$ of subsets $A, B \subseteq X$ such that $A$ is the set of lower bounds for $B$ and $B$ is the set of upper bounds of $A$. For the rationals $\mathbb{Q}$, these pairs correspond to Dedekind cuts, and the completion of $\mathbb{Q}$ as a lattice is the lattice of the reals $\mathbb{R}$.
\end{remark}

@@

\begin{defn}[Boolean algebra]
	An algebraic \marginleft{Boolean \\ algebra} structure $(X, \vee, \wedge, \neg, \bot, \top)$ is a \textbf{Boolean algebra} if it satisfies the following properties:
	\begin{description}
		\item[Lattice] $(X, \vee, \wedge)$ is a lattice
		\item[Distributivity] $a \vee (b \wedge c) = (a \vee b) \wedge (a \vee c)$
and $a \wedge (b \vee c) = (a \wedge b) \vee (a \wedge c)$
		\item[Identity] $a \vee \bot = a$
		and $a \wedge \top = a$
		\item[Complementation] $a \vee \neg a = \top$
		and $a \wedge \neg a = \bot$
	\end{description}
\end{defn}

@@

\begin{prop}[Stone's representation theorem]
	Every bounded distributive lattice is order isomorphic to a lattice of subsets ordered by inclusion.
\end{prop}
